---
title: "Dispute Resolution: Exploratory Analysis"
output: html_notebook
---

# Purpose and Initial Questions 

The purpose of this notebook is to clean the federal dispute resolution data so it lends itself a bit better to understanding, exploration, analysis. Some 
questions to explore are: 

- Are there any interesting patterns in variation in any of the variables?  
- What proportion of each dispute total category does each state represent?  
- Is there a relationship betweent the population of a state and the number of dispute totals?  

```{r}
library(tidyverse, verbose = T)
library(stringr, verbose = T)
library(maps, verbose = T)
```

First, we'll read in the data and combine the 2012 through 2015 datasets into one dataframe. For now, we won't `gather` up the variables so that we can explore the variation in each variable. 

```{r read files}
fnames <- c(
  "https://www2.ed.gov/programs/osepidea/618-data/state-level-data-files/part-b-data/dispute-resolution/bdispres2014-15.csv",
  "https://www2.ed.gov/programs/osepidea/618-data/state-level-data-files/part-b-data/dispute-resolution/bdispres2013-14.csv", 
  "https://www2.ed.gov/programs/osepidea/618-data/state-level-data-files/part-b-data/dispute-resolution/bdispres2012-13.csv")
  
# TODO: Work on reading these CSVs
# "https://www2.ed.gov/programs/osepidea/618-data/state-level-data-files/part-b-data/dispute-resolution/bdispres2011-12.csv",
# "https://www2.ed.gov/programs/osepidea/618-data/state-level-data-files/part-b-data/dispute-resolution/bdispres2010-11.csv"

# Read files
allfiles <- lapply(fnames, function(x) {
  read_csv(x, skip = 4, col_types = cols(Year = col_character()), n_max = 61)
  })

# Combine files
dpr <- bind_rows(allfiles)
```

Let's rename the variables to something manageable but still descriptive. 
We'll use the following conventions:  

- Variables prefixed with `wsc` refer to categories that related to written and signed complaints.  
- Variables prefixed with `med` refer to categories related to mediations.  
- Variables prefixed with `edpc` refer to categories related to expedited due process complaints.  

```{r}
dpr <- dpr %>% 
  rename(# Written signed complaints
         wsc_total = `Written, Signed Complaints (WSC) Total (1)`, 
         wsc_reports = `WSC with Reports Issued Total (1.1)`, 
         wsc_reports_with_findings = `WSC Reports with Findings (1.1a)`, 
         wsc_reports_ontime  = `WSC Reports within Timeline (1.1b)`, 
         wsc_reports_ext_timelines = `WSC Reports within Extended Timelines (1.1c)`, 
         wsc_pending = `WSC Pending Total  (1.2)`, 
         wsc_pending_due_proc = `WSC Pending a Due Process Hearing (1.2a)`, 
         wsc_withdrawn = `WSC Withdrawn or Dismissed(1.3)`, 
         # Mediations
         med_total = `Mediation Requests Total (2)`, 
         med_held = `Mediations Held Total (2.1)`, 
         med_from_due_proc = `Mediations Held Related to Due Process Complaints(2.1a)`, 
         med_agreed_from_due_proc = `Mediation Agreements Related to Due Process Complaints (2.1ai)`, 
         med_not_from_due_proc = `Mediations Held Not Related to Due Process Complaints (2.1b)`,
         med_agreed_not_from_due_proc = `Mediation Agreements Not Related to Due Process Complaints (2.1bi)`, 
         med_pending = `Mediations Pending (2.2)`, 
         med_withdrawn = `Mediations Withdrawn or Not Held (2.3)`, 
         # Due process
         dpc_total = `Due Process Complaints (DPC) Total (3)`, 
         dpc_resolutions = `DPC Resolution Meetings Total (3.1)`, 
         dpc_resolutions_settlements = `DPC Resolution Meetings - Written Settlement Agreements (3.1a)`, 
         dpc_adjudicated = `DPC Hearings (fully adjudicated) Total (3.2)`, 
         dpc_decisions_ontime = `DPC Written Decisions within Timeline (3.2a)`, 
         dpc_decisions_ext_timeline = `DPC Written Decisions within Extended Timelines (3.2b)`, 
         dpc_pending = `DPC Pending (3.3)`, 
         dpc_withdrawn = `DPC Withdrawn or Dismissed (3.4)`, 
         # Expedited due process complaints
         edpc_total = `Expedited Due Process Complaints (EDPC) Total (4)`, 
         edpc_resolutions = `EDPC Resulted in a Resolution Meeting Total (4.1)`, 
         edpc_resolutions_settlements = `EDPC Resolution Meetings - Written Settlement Agreements (4.1a)`, 
         edpc_adjudicated = `EDPC Expedited Hearings (fully adjudicated) Total (4.2)`, 
         edpc_placement_changes = `EDPC Expedited Hearings - Change of Placement Ordered (4.2a)`, 
         edpc_pending = `EDPC Pending (4.3)`, 
         edpc_withdrawn = `EDPC Withdrawn or Dismissed (4.4)`)
```

```{r}
# TODO: Create a new categorical variable for the type of process 
# (e.g., EDC, mediations, etc.)
```

We'll also clean up our state names to all lower case to make it easier to plot our map later in the analysis. 

```{r}
dpr$State <- tolower(dpr$State)
```

Let's verify that we have usable classes for each of the variables. 

```{r}
str(dpr)
```

```{r}
# Write csv
write_csv(dpr, path = "output/2012_2015_dispute_resolution.csv")
```

```{r}
# Clean up unneeded variables 
rm(list = c("allfiles", "fnames"))
```

# Exploring the Data 

Here are the counts for the number of observations in each year: 

```{r}
count(dpr, Year)
```

There are some really strange values, which might indicate some data errors on 
the high end. We can illustrate that here in the 2014-2015 data. 

```{r}
dpr1415 <- dpr %>% 
  filter(Year == "2014-2015")

# Compute the range of each variable
sapply(dpr1415[, 3:length(names(dpr1415))], function(x) {range(x)})
```

Let's further explore the strange high values in some of the categories. The 
outliers are so high and so different from the the rest of the value that it 
obscures the boxplots so we're going to zoom in to `total` values of 5000 or less.

For most categories, the median count across all states do not give a good summary of any one state's totals. Most categories have states whose totals are uncharacteristically high for that category. In other words, the categories 
are characterized by low medians with some states containing noticeable 
outliers. 

```{r}
# Gather variables
dpr_narrow <- dpr %>% 
  gather(category, total, -c(Year, State))
```

```{r}
# TODO: Label max values for each category with the state name
ggplot(data = dpr_narrow) +
  geom_boxplot(aes(x = reorder(category, total, FUN = max), y = total), 
               alpha = .5, 
               color = "cyan4") + 
  coord_flip(ylim = c(0, 5000)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Distribution of Category Totals", 
       subtitle = "Data: State level data from ed.gov", 
       x = "", 
       y = "") + 
  facet_wrap(~Year)
```

# How are these totals related to each state's population? 

```{r}
state_pop <- read_csv("state_data/state_pop.csv")
state_pop$State <- tolower(state_pop$State) # Change state names to lower case

# Join the state population dataset to the dispute resolution dataset
dpr_pop <- left_join(dpr_narrow, state_pop, by = c("State" = "State"))
```

I do not have population data for the following regions, so for now we will leave them out and will examine them later. In particular, US, Outlying Areas and Freely Associated States has a sizable amount of category totals.

```{r}
unique(dpr_pop$State[is.na(dpr_pop$Population)])
```

```{r}
# Remove areas that do not have population values 
dpr_pop <- dpr_pop %>% 
  filter(!is.na(Population))
```

Note here that not all categories suggest a relationship between a state's population and the total count of that category. 

```{r}
ggplot(data = dpr_pop) + 
  geom_point(aes(x = Population, y = total), na.rm = T, alpha = .5) + 
  geom_smooth(aes(x = Population, y = total)) +
  coord_cartesian(ylim = c(0, 1000)) +
  facet_wrap(~ category) + 
  labs(title = "State Population and Category Counts", 
       subtitle = "Data: State level data from ed.gov")
```

```{r}
# TODO: Regress `total` on `Population` for categories with a 
# possible relationship
```

# Visualizing 2014-2015 Totals on a US Map

```{r}
# Subset the 14-15 data and filter for totals categories 
dpr_narrow_1415 <- dpr_pop %>% 
  filter(Year == "2014-2015", 
         category %in% c("wsc_total", "med_total", "dpc_total", "edpc_total")) %>% 
  mutate(category = ifelse(category == "wsc_total", "Written Complaints", category), 
         category = ifelse(category == "med_total", "Mediations", category), 
         category = ifelse(category == "dpc_total", "Due Process", category), 
         category = ifelse(category == "edpc_total", "Expedited Due Process", category), 
         per_capita = 100 * (total / Population))
```

```{r}
# TODO: Map fill to per capita total
all_states <- map_data("state") 

# Merge the states data with dpr data
dpr_map <- left_join(dpr_narrow_1415, all_states, by = c("State" = "region"))

# Clean data to prepare it for the map plot
dpr_map <- dpr_map %>% 
  filter(!is.na(long)) %>%
  arrange(group, order)
```

Here are how category totals are distributed across states. 

```{r}
# Total counts
ggplot(data = dpr_map, 
       aes(x = long, y = lat, group = group, fill = total)) +
  geom_polygon(colour = "white") + 
  labs(title = "Total Count of 2014-2015 Dispute Events", 
       subtitle = "Data: State level data from ed.gov", 
       fill = "", 
       x = "", 
       y = "") +
  facet_wrap(~category)
```

```{r}
# Total counts per capita
ggplot(data = dpr_map, 
       aes(x = long, y = lat, group = group, fill = per_capita)) +
  geom_polygon(colour = "white") + 
  labs(title = "2014-2015 Dispute Events Per Capita", 
       subtitle = "Data: State level data from ed.gov", 
       fill = "", 
       x = "", 
       y = "") +
  facet_wrap(~category)
```

First let's plot the variation in each variable.

```{r}
ggplot(data = dpr_pop, aes(x = total)) + 
  geom_histogram(binwidth = 15) +
  facet_wrap(~ category)
```

```{r by state in 14-15}
# Total due process complains in 14-15 by state\
dpc1415 <- dpr %>% 
  select(Year, State, dpc_total) %>% 
  filter(Year == "2014-2015") %>%
  filter(min_rank((desc(dpc_total))) < 10)

ggplot(data = dpc1415, aes(reorder(x = State, dpc_total), y = dpc_total)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Total Due Process Complaints", 
       subtitle = "2014-2015", 
       x = "", y = "Total Due Process Hearings")
```

```{r by state in 13-14}
# TODO: Include other years and spread by year
# Total due process complains in 13-14 by state
by_state_13 <- dpr %>%
  filter(Year == "2013-2014") %>% 
  select(Year, State, dpc_total) %>% 
  arrange(desc(dpc_total))
```

```{r}
# What is the median number of due process complaints?
p <- ggplot(data = dpr, aes(x = Year, y = dpc_total)) +
  geom_boxplot() + 
  labs(title = "Median Total Due Process Complaints", 
       subtitle = "Data: Federal Dispute Resolution Data", 
       y = "")

# Zoom in on totals of less than 500
p + coord_cartesian(ylim = c(0, 500))
```

```{r}
# TODO Need to find a better way to explore year to year
#ggplot(data = filter(
#  dpr,
#  category == "Due Process Complaints (DPC) Total (3)"),
#  aes(x = State, y = total)
#  ) +
#  geom_bar(stat = "identity") +
#  coord_flip() +
#  facet_wrap(~ Year)
```

```{r mediation agreements}
# What is the relationship between mediations held and mediation agreements?
# This looks just at mediations relating to due process hearings.
med_plot <- ggplot(data = dpr,
                   aes(x = med_from_due_proc, y = med_agreed_from_due_proc)) +
  geom_point(position = "jitter", alpha = .50) + 
  geom_smooth() +
  facet_wrap(~ Year) +
  labs(title    = "Mediations Held vs. Mediation Agreements",
       subtitle = "Dataset includes only mediations related to due process complaints.",
       x = "Mediations held relating to due process", 
       y = "Agreements reached")

# Zoom in on x < 1000 and 6 < 500
med_plot + coord_cartesian(xlim = c(0, 1000), ylim = c(0, 500))
```